---
title: "Cleaning Data"
author: "Varun Boodram"
date: "September 24, 2014"
output: html_document
---
Steps in obtaining data
1. Create a directory to store the file in 
2. Download the file with download.file()
3. read in the file

*read.table() is the most commonly used 
```{r, cache=T}
getandload<-function(x){
        if(!file.exists("./restaurantdata")){
                dir.create(path = "./restaurantdata")
        }
        fileurl<-x
        if(!file.exists("./restaurantdata/restaurant.csv")){
                download.file(url = fileurl, destfile = "./restaurantdata/restaurant.csv", method = "curl") #destfile is what we want to call the downloded file. The extension reflects the type of file
                datedownloaded<-date()
        }
        read.csv("./restaurantdata/restaurant.csv", header = T, sep = ",")
}
fileUrl<-"https://data.baltimorecity.gov/api/views/k5ry-ef3g/rows.csv?accessType=DOWNLOAD"
resdat<-getandload(fileUrl)
list.files("./restaurantdata")
```
The key process of data cleaning is looking at the data that you've loaded into R, and identifying any missing values, or other quirks or weird issues that you need to address before you do downstream analysis.   
eg: data from [Open Baltimore](https://data.baltimorecity.gov/Culture-Arts/Restaurants/k5ry-ef3g), Restaurant Data

Steps in cleaning data

1. Create data summaries

head() and tail()
```{r}
list(head=head(resdat, 3), tail=tail(resdat,3))
```
summary() gives an overall summary of the data. For every single varible, it will return some information. For text based or factor variables, it will give the count of each factor. Below, there are 8 McDonald's, 7 Popeyes, and 6 Subways. Quantitave data is given a five-number summary. 
```{r}
summary(resdat)
```
str() will give some more info about the data frame, such as class, dimension, and the names of each column, and what class of vector each column is. Below, name is a factor vector with 1277 levels
```{r}
str(resdat)
```
quantile() will return the variablity of quantitative variables. 
```{r}
quantile(resdat$councilDistrict, na.rm = T)
```
table() tabulates the data for specific data, useNa will tell you how many missing values there may be. Two dimensional tables can give a sense of the relationships between variables, 

```{r}
table(resdat$zipCode, useNA = "ifany")
head(table(resdat$councilDistrict, resdat$zipCode), 3)
# find values with specific characteristics
table(resdat$zipCode %in% "21212") # return all variable 1 that are in variable 2
table(resdat$zipCode %in% c("21212", "21213"))
```
As an aside, we can subset with the logical vector 
```{r}
head(resdat$zipCode %in% c("21212", "21213"))
head(resdat[resdat$zipCode %in% c("21212", "21213"), ], 3)
```
xtabs() allows for cross tabulation, a statistical process that summarizes categorical data to create a contingency table. 
```{r}
data(UCBAdmissions)
df<-as.data.frame(UCBAdmissions)
df
summary(df)
#xtabs(col you want to examine ~ broken down by + broken down by +.., data)
xtabs(formula = Freq~Gender+Admit, data = df)
data(warpbreaks); head(warpbreaks, 10)
xtabs(breaks~., data = warpbreaks) #breaks broken down by all other variables in the data set
```
The summary not as complete as it can be. The [help page](http://www.inside-r.org/r-doc/datasets/warpbreaks) for warpbreaks says, "There are measurements on 9 looms for each of the six types of warp", but these 9 looms are not represented in the xtabs. Conversion to a flat table remedies this
```{r}
warpbreaks$loom<-rep(1:9, nrow(warpbreaks)/9)
head(warpbreaks)
xt<-xtabs(breaks~., data = warpbreaks); xt
ftable(xt)
