---
title: "Reading data from databases"
author: "Varun Boodram"
date: "October 15, 2014"
output: html_document
---

### mySQL (pr: my sequel)

mySQL is an open-source database, that is widely used in internet applications. The data are structured in databases, which nest connected tables, which nest fields. Each row is a record.   
eg: a table of employees may be linked above to their managers, which is linked to *their* departments; the table of employees may also be linked to a table of their titles, which is linked to a table of their salaries.
The take-away is that each table represents a different data type; each table corresponds to a data-frame in R.   
Each time you perform a **query**, you must use one of the commands from the mySQL language. Common ones ([full list](http://www.pantz.org/software/mysql/mysqlcommands.html) include
Create a database on the sql server.
```
mysql> create database [databasename];
```
List all databases on the sql server.
```
mysql> show databases;
```
Switch to a database.
```
mysql> use [db name];
```
To see all the tables in the db.
```
mysql> show tables;
```
To see database's field formats.
```
mysql> describe [table name];
```
To delete a db.
```
mysql> drop database [database name];
```
To delete a table.
```
mysql> drop table [table name];
```
Show all data in a table.
```
mysql> SELECT * FROM [table name];
```
Returns the columns and column information pertaining to the designated table.
```
mysql> show columns from [table name];
```
Show certain selected rows with the value "whatever".
```
mysql> SELECT * FROM [table name] WHERE [field name] = "whatever";
```
Show all records containing the name "Bob" AND the phone number '3444444'.
```
mysql> SELECT * FROM [table name] WHERE name = "Bob" AND phone_number = '3444444';
```
Show all records not containing the name "Bob" AND the phone number '3444444' order by the phone_number field.
```
mysql> SELECT * FROM [table name] WHERE name != "Bob" AND phone_number = '3444444' order by phone_number;
```
Show all records starting with the letters 'bob' AND the phone number '3444444'.
```
mysql> SELECT * FROM [table name] WHERE name like "Bob%" AND phone_number = '3444444';
```
Show all records starting with the letters 'bob' AND the phone number '3444444' limit to records 1 through 5.
```
mysql> SELECT * FROM [table name] WHERE name like "Bob%" AND phone_number = '3444444' limit 1,5;
```
Use a regular expression to find records. Use "REGEXP BINARY" to force case-sensitivity. This finds any record beginning with a.
```
mysql> SELECT * FROM [table name] WHERE rec RLIKE "^a";
```
Show unique records.
```
mysql> SELECT DISTINCT [column name] FROM [table name];
```
Show selected records sorted in an ascending (asc) or descending (desc).
```
mysql> SELECT [col1],[col2] FROM [table name] ORDER BY [col2] DESC;
```
Return number of rows.
```
mysql> SELECT COUNT(*) FROM [table name];
```
Sum column.
```
mysql> SELECT SUM(*) FROM [table name];
```
Join tables on common columns.
```
mysql> select lookup.illustrationid, lookup.personid,person.birthday from lookup left join person on lookup.personid=person.personid=statement to join birthday in person table with primary illustration id;
```
```{r, message=FALSE}
library(RMySQL)
```
A major role of the data scientist is, of course, obtaining data from a database, and putting data back into it. Obtaining data is the more common task.   
eg: Accessing the [UCSC Human Genome database](https://genome.ucsc.edu/) to obtain information on a particular genome we are interested in. Go to downloads>>mySQL access for information on how to connect

1. Connect to the database with dbConnect()

* each connection should be assigned a handle
```{r}
# user and host info is found on the access page
ucscdb<-dbConnect(drv = MySQL(), 
                  user="genome", 
                  host="genome-mysql.cse.ucsc.edu")
# Check the above list for the command to list all databases on the server. This is taken as an argument of dbGetQuerry()
result<-dbGetQuery(conn = ucscdb,  
                   statement = "show databases;")
# it is good practise to ***always*** disconnect from the server at the end of your querry
dbDisconnect(conn = ucscdb)
head(result, 10)
```

2. Connecting to a particular database  

hg19 is a particular build of the human genome that we are interested in. It is a **database**, and what we are interested in is listing the **tables** that it contains

```{r}
# assign a handle, and also include the database name in dbConnect()
hg19<-dbConnect(drv = MySQL(), 
                  user="genome", 
                  db = "hg19", 
                  host="genome-mysql.cse.ucsc.edu")
# there is a mySQl statement to see all the tables in a db (check the list above), but this can also (and prefferentially) be accomplished with the RMySQL command dbListTables()
allTables<-dbListTables(conn = hg19)
dbDisconnect(hg19)
```

Remember that each table is like a data frame in R. 
```{r}
allTables[1:20]
which(allTables %in% "affyU133Plus2")
```

3. Getting information from a particular table in a database

```{r}
# because the connection was previously closed, we need to reopen it
hg19<-dbConnect(drv = MySQL(), 
                  user="genome", 
                  db = "hg19", 
                  host="genome-mysql.cse.ucsc.edu")
#list fields 
dbListFields(conn = hg19, 
             name = "affyU133Plus2")
# find the command to look up the number of rows in a dataset in the above list. perform a query to get that information
dbGetQuery(conn = hg19, 
           statement = "SELECT COUNT(*) FROM affyU133Plus2;")
```

4. Store a table locally
The following obtains a table from a database, but you shold be warned that the amount of data stored is usually pretty huge. This is both time and memory consuming
```{r, cache=TRUE}
if(!exists("affyData")){
        affyData<-dbReadTable(conn = hg19, name = "affyU133Plus2")
}
head(affyData)
```
It is far more useful to select a particular subset of the table for analysis whenever you can. 
```{r}
# find the command to select a subset of rows from a table in the above list
query<-dbSendQuery(conn = hg19, 
                  statement = "SELECT * FROM affyU133Plus2 WHERE misMatches between 1 and 3;")
#use fetch() to get the results of the query
affyMis<-fetch(query)
dim(affyMis)
# get a subset of affyMis
smallAffyMis<-fetch(res = query,
                    n = 10 )
dim(smallAffyMis)
# always clear the querry after you are done
dbClearResult(res = query)
dbDisconnect(hg19)
```

### HDF5
Heirarchical Data Format have 

* groups containing data and metadata
Each group has a group header, and a group symbol table that has a list of objects in the group
* datasets
With header, datatype, dataspace, and storage layout, and a data array (a data frame)
```{r, message=FALSE}
library(rhdf5)
```
Here we create hdf5 files and interact with them
```{r}
# create file (ext .h5)
created<-h5createFile(file = "example.h5")
created
# create groups within the files
created<-h5createGroup(file = "example.h5", group = "foo")
#create a subgroup of foo
created<-h5createGroup(file = "example.h5", group = "foo/foobar")
created<-h5createGroup(file = "example.h5", group = "bar")
# view groups
h5ls(file = "example.h5")
# write data to specific groups
A<-matrix(1:10, nrow = 5, ncol = 2)
# write the object to a group within the file
h5write(obj = A, file = "example.h5", name = "foo/A")
h5ls(file = "example.h5")
B<-array(data = seq(from = 0.1, to = 2, by = 0.1), dim = c(5,2,2))
B
h5write(obj = B, file = "example.h5", name = "foo/foobar/B")
h5ls(file = "example.h5")
# write a dataset to the top level group
df<-data.frame(1L:5L, seq(from = 0, to = 1, length.out = 5), row.names = c("ab", "cde","fgh", "a", "s"), stringsAsFactors = F)
df
h5write(obj = df, file = "example.h5", name = "df")
h5ls(file = "example.h5")
# read data with h5read()
h5read(file = "example.h5", name = "foo/A")
h5read(file = "example.h5", name = "foo/foobar/B")
#expanding data sets by reading and writing chunks
# Wrtie stuff to a specific part of A
h5write(obj = c(12,13,14), file = "example.h5", name = "foo/A", index=list(1:3, 1))
h5read(file = "example.h5", name = "foo/A")
